---
title: "NHL xG"
output:
  github_document:
    toc: true
  code_folding: show
  toc-title: "Table of Contents"
  number_sections: true
#  toc_depth: 2
#    code_folding: show
#    mathjax: local
#    self_contained: false
#header-includes: -\usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2023)
library(sportyR)
library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)
library(knitr)
require(tidymodels)
library(readr)
library(rmarkdown)
library(pROC)
library(randomForest)
library(cowplot)
library(usethis)
library(parallel)
library(Hmisc)
library(glmnet)
library(mikropml)
```

## Overview and Scope

The aim of this project is to build different naive Expected Goals models using LPM, logit and possibly other regression models at some point. The data is obtained from [Kaggle](https://www.kaggle.com/datasets/martinellis/nhl-game-data) and valuables sources that have been utilized are in this [section](Sources).

Unfortunately Github-flavored documents don't support code-folding, and hence as of now the notebook is showing all the code even for the plots. 

## Data Wrangling

Loading the data and creating a variable for shot data

```{r load_data, cache = TRUE, R.options = list(width = 10000)}
game_plays <- read.csv("~/Downloads/archive/game_plays.csv")

## Making sure the data looks correct, showing first 18 samples and the response variable
str(game_plays)
```

Parsing the data to include only the shots and goals

```{r shots, cache=TRUE}
shots <- na.omit(game_plays) %>%
  filter(event %in% c("Goal", "Shot")) %>%
  select(team_id_for, event, st_x, st_y, secondaryType)

## Creating a new column for a binary response variable if a shot was a goal or not
shots$goal <- as.integer(ifelse(shots$event == "Goal", 1,0))
```

Creating local variables for x and y cordinates

```{r cordinates}

##Taking the absolute value to have all the shots in the same horizontal axis
x <- abs(shots$st_x)
y <- shots$st_y
```

Creating functions for distance and angle. Distance from the shot place to the center of the goal is calculated by

$c^2 = a^2 + b^2 \Rightarrow c = \sqrt{a^2 + b^2 }$

The angle visible to the player is calculated as follows,

$\tan \theta = \frac{\text{goal width } * x }{x^2 + y^2 - (\frac{\text{goal width}}{2})^2 }$

```{r functions}
distance <- function(x_pos, y_pos) {
  sqrt((89 - abs(x_pos))^2 + y_pos^2)
}

angle_theta <- function(x_pos, y_pos) {
  x_temp <- abs(89 - abs(x_pos))
    ifelse((x_temp)^2 + (y_pos)^2 >= 9, atan((6 * abs(x_temp))/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi,
    180 + atan((6 * x_temp)/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi)
}

## Making sure our range is correct, should be between 0 and 180
range(angle_theta(x,y),na.rm=TRUE)

## adding columns for distance and angle
shots <- shots %>%
  mutate(distance = distance(shots$st_x, shots$st_y),
                    angle = angle_theta(shots$st_x, shots$st_y))
## Testing this works
head(shots)
```

## Data Exploration

### Histograms

a couple of histograms from the data that show the distribution of shot distance and angle

```{r, Histograms, fig.show="hold", out.width="50%", class.source = 'fold-hide'}
par(mar = c(4, 4, .1, .1))
ggplot(shots, aes(x=shots$angle)) +
  geom_histogram(binwidth = 3,
                 center = 0,
                 color = "black",
                 fill = "white") +
  scale_x_continuous(limits = c(0, 180)) +
  theme_bw() +
  theme(panel.border = element_blank()) + 
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank()) + 
  xlab("angles") + 
  ylab(element_blank()) + 
  ggtitle("Histogram of Shot angles")

ggplot(shots, aes(x=distance)) +
    geom_histogram(binwidth = 6,
                   center = 0,
                   color = "black",
                   fill = "white") +
    scale_x_continuous(limits = c(0, 180)) +
    theme_bw() +
  theme(panel.border = element_blank()) + 
    theme(panel.grid.major = element_blank()) +
    theme(panel.grid.minor = element_blank()) + 
    xlab("Distance") + 
    ylab(element_blank()) + 
    ggtitle("Histogram of Shot Distances")
```

### Probability of a Goal Given Distance or angle

```{r Bins, fig.show="hold", out.width="50%", cache = TRUE}

bins_distance <- aggregate(shots,
                   by=list(cut(shots$distance, seq(0,100,10))),
                   mean)

bins_angle <- aggregate(shots,
                  by=list(cut(shots$angle, seq(0,180,10))),
                  mean)

## Changing the first column to numeric values so that ggplot geom_smooth works
bins_distance$Group.1 <- as.numeric(bins_distance$Group.1)
bins_angle$Group.1 <- as.numeric(bins_angle$Group.1)

angles <- as.character(seq(0, 180, 10))
distances <- as.character(seq(0, 90, 10))


ggplot(bins_distance, aes(x= bins_distance$Group.1, y =  bins_distance$goal)) +
                          geom_point() +
                          geom_line() +
                          theme_bw() + 
                          xlab("Distance to goal (Feet)") +
                          ylab("Probability of Goal") + 
                          scale_x_discrete(limits = distances) +
                          ggtitle("Probability of Goal Given the Distance")
                          
ggplot () + aes(x= bins_angle$Group.1, y =  bins_angle$goal) +
  geom_point() +
  geom_smooth(method=lm, se = F) + 
  theme_bw() +
  xlab("angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Probability of Goal Given the angle") +
  scale_x_discrete(limits = angles)
```

In the distance to goal there's an interesting fact: probability of goal increases with distance. This is likely due to the fact that usually shots from very far away are shot due to empty goal: hence it scewing the data. In angle to goal there's no notable surprises.


### Shot Plots
```{r shotPlots, fig.show="hold", out.width="50%"}
#Shot maps for randomly samples 2500 shots for faster computations and more informative plots 
shots_parsed <- shots %>%
  subset(st_x < 88) %>%
  sample_n(2500)

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = shots_parsed$st_y, y = shots_parsed$st_x, col = shots_parsed$goal, alpha = 0.9, size = shots_parsed$goal)) +
  scale_color_binned(low ="red", high = "darkgreen") +
  scale_size_continuous(range = c(1,3))

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_density2d_filled(aes(x = shots_parsed$st_y, y = shots_parsed$st_x, alpha = 0.5),
                        contour_var = "ndensity",
                        breaks = seq(0.1, 1.0, length.out = 10))
```


## Model Building

Here we split the data as a cross validation technique, using 70% of the data as the training data and the rest 30% as the testing data. A new explanatory variable - secondaryType - is introduced, which includes information on the shot type. It will be used in the later models. 

```{r training_data}
unique(shots$secondaryType)

## Checking for NA values 
sum(is.na(shots))

parsed_shots <- na.omit(shots) %>%
  select(goal, distance, angle, secondaryType) %>%
  mutate(distance_sq = (distance^2))
  
train_test_split <- initial_split(data = parsed_shots, prop = 0.7)

train_data <- train_test_split %>%
  training()
test_data <- train_test_split %>%
  testing()
```
Because the quantity of NA's was so small, we can just omit them from the further analysis.

### Linear Probability Model

```{r LPM}
LPM <- lm(goal ~ distance + angle, data = train_data)
summary(LPM)
```

In the plot below, the the main downside of LPM model becomes apparent: results are not bound [0,1].


```{r Heatmap LPM, fig.show="hold", out.width="50%"}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$distance_sq <- artificial_shots$distance^2
artificial_shots$xg <- predict(LPM, artificial_shots, type = "response")

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg, alpha = 1)) +
  scale_color_gradient2(low = "white", mid="red", midpoint = 0.55, high ="darkred",
                       scales::rescale(c(0.9,0.1))) 

ggplot(data = LPM, mapping=aes(x=angle, y = goal)) +
  geom_point(size = 0.01) + geom_smooth(method = "lm", se = F, size = 0.5) +
  theme_bw()
```

### Logit
Due to the significant downsides of LPM, logistic regression is henceforth used. 

```{r logit}
logit <- glm(goal ~ distance + angle,
             family = binomial(link = 'logit'),
             data = train_data)
```

In a logit model, the probability of an event is given by

$P = \frac{1}{1 + - exp^{-{(\beta_0 + \beta_1 x_1 \beta_2 x_2 + …)}}}$

```{r Logit Plots, fig.show="hold", out.width="50%"}
ggplot(logit, aes(x=distance, y =goal)) +
  geom_point() + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  theme_bw() +
  xlab("Distance to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Distance as an explanatory variable") 

ggplot(logit, aes(x=angle, y =goal)) +
  geom_point(size = 0.01) + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  theme_bw() +
  xlab("angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("angle as an explanatory variable") 
```

From graphs above, it becomes visually clear that angle is a way more important factor affecting if a shot is a goal or not. To test whether we could improve explanatory power of distance, we add a quadratic form of it as an extra variable.

```{r Quadratics data for logit}
logit.2 <- glm(goal ~ distance + distance_sq + angle,
               family = binomial(link = 'logit'),
               data = train_data)

logit_coef <- logit$coefficients
logit_distance <- logit_coef["distance"]
logit_intercept <- logit_coef["(Intercept)"]

logit.2_coef <- logit.2$coefficients
logit.2_distance <- logit.2_coef["distance"]
logit.2_distance_sq <- logit.2_coef["distance_sq"]
logit.2_intercept <- logit.2_coef["(Intercept)"]

b <- data.frame(c(seq(0,100,.1)))
a <- (1 / (1 + exp(-logit.2_distance * b - logit.2_distance_sq * b - logit.2_intercept)))
a.2 <- (1 / (1 + exp(-logit_distance * b - logit_intercept)))
c <- cbind(a, a.2, b)

colnames(c) <- c("a", "a.2", "b")
```


```{r Heatmap data for logit}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$distance_sq <- artificial_shots$distance^2

artificial_shots$xg_logit <- predict(logit.2, artificial_shots, type = "response")
```

```{r LogitPlots1, fig.show="hold", out.width="50%"}
ggplot(c, aes(x=b,y=a)) +
  geom_point(size = 0.1, col = "darkgreen") + 
  geom_point(aes(y=a.2), size = 0.1, col = "darkred") +
  theme_bw() + 
  ggtitle("Comparing Distance Variables with and without quadratic term") +
  xlab("Distance to Goal") +
  ylab("Probability of a Goal")

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg_logit, alpha = 0.1)) +
  scale_color_gradient(low = "white", high ="red",
                       scales::rescale(c(0.1,0.9))) 
```

As we can see from the graph above, including the quadratic term makes a difference but not a suitable one. Namely, it decreases the probability when close the goal and increases it when further away, increasing the 'empty-net bias'. Hence, we keep using the original logit model as the baseline. 



Next, we add the previously mentioned shot type into the regression and compare the perfomance between that and the baseline model. 

```{r adding secondaryType}
logit.3 <- glm(goal ~ distance + angle + secondaryType,
             family = binomial(link = 'logit'),
             data = train_data)

#Creating vectors of predictors for the new model with secondaryType and for the last one without it 
logit_pred <- predict(logit, test_data, type = "response")
logit.3_pred <- predict(logit.3, test_data, type = "response")

#Comparing Logit models with AUC values
roc.test(roc(test_data$goal, logit.3_pred), roc(test_data$goal, logit_pred))

# Comparing models with AIC values 
print(paste("AIC without shot type:", logit$aic))
print(paste("AIC with shot type:", logit.3$aic))

par(pty = "s")

logit.3_roc <- roc(test_data$goal, logit.3_pred, plot = TRUE, print.auc = TRUE, col = "darkred",
  legacy.axes = TRUE, percent = TRUE, xlab = "False Positive Percentage",
  ylab = "True Positive Percentage") 


#Observing the optimal threshold level and the corresponding specificity and sensitivity levels
logit.3_threshold <- coords(logit.3_roc, "best", "threshold")
logit.3_threshold
```

Here is one of the puzzles when using a logit model: which threshold to use when categorizing a shot. Using the coords function, we find the threshold which maximizes both accuracy and specifity, but not the model accuracy. After trying different thresholds, we could get the model accuracy above 90%, but with the cost that specifity drops down to under 30%. Therefore, for the purposes of this project we continue with the value given by the coords function. 

```{r testing logit}
#Creating Confusion Matrix
logit.3_conf <- table(logit.3_pred>=logit.3_threshold$threshold, test_data$goal)
logit.3_conf

#Model accuracy
sum(diag(logit.3_conf))/sum(logit.3_conf)*100

## Calculating Pseudo R^2
ll.null <- logit.3$null.deviance/-2
ll.proposed <- logit.3$deviance/-2

(ll.null - ll.proposed) / ll.null

## Calculating the p-value for the pseudo R squared
1 - pchisq(2*(ll.proposed - ll.null), df = (length(logit.3$coefficients) -1))
```

The pseudo r-squared is very low and the p-value is zero, making the statistic robust. 

```{r logit overview plot }
pred_logit.3 <- data.frame(prob.of.goal = logit.3$fitted.values,
                           goal = train_data$goal)

pred_logit.3 <- pred_logit.3[
  order(pred_logit.3$prob.of.goal, decreasing = FALSE),]

pred_logit.3$rank <- 1:nrow(pred_logit.3)

ggplot(data = pred_logit.3, aes(x = rank, y = prob.of.goal)) +
  geom_point(aes(color = goal), alpha = 1, shape = 4, stroke = 2) + 
  scale_color_binned(low = "red", high = "navyblue") +
  theme_bw() +
  labs(x = "Index",
       y = "Probability of a Goal",
       title = "Overview of the Logit Model")

```

Comparing this analysis to the other similar ones, people usually stop at this point because for such simple binary classification, logit if usually enough. There's no sense in applying more advanced models if we can get by with this. Moreover, tuning hyper parameters (L1, L2, elastic-net) similarly have empirically made very little difference to the results with so little explanatory variables. 

However, I will start tuning the parameters for the logit model and introduce other machine learning classification models for own educational purposes. Further analysis will be done with the package 'mikropml' (which is built on top of 'caret') when adjusting the L2 Ridge regularization parameter, as elastic net and L1 in this case would be useless due of the importance of all variables. Pre-processing steps to be made to prepare data for the packages mentioned: 

* Creating own column for each shot type and removing the original column
* Modifying outcome column to not have spaces in the end result 

Data will not be centered (at least at this point) to save interpretability. Missing values are not an issue, as we omitted all 28 of them in the beginning. 

This dataset is very small compared to the full sample (~900k shots) due to the heavy computation made by caret and mikropml. This will possibly be solved in the future by using parallelized processing. 

```{r, cache = TRUE}
library(caret)
library(mikropml)

parsed_shots.logit <- parsed_shots %>%
  select(goal, distance, angle, secondaryType) %>%
  sample_n(size = 2500)

dummy_shots <- dummyVars("~.", data = parsed_shots.logit)
parsed_shots.logit <- data.frame(predict(dummy_shots, newdata = parsed_shots.logit))

parsed_shots.logit$goal <- ifelse(test=parsed_shots.logit$goal == "1", yes = "goal", no = "notagoal")
parsed_shots.logit$goal <- as.factor(parsed_shots.logit$goal)
str(parsed_shots.logit)

get_srn_results <- run_ml(parsed_shots.logit,
                          method ="glmnet",
                          outcome_colname = "goal",
                          training_frac = 0.8,
                          kfold = 3,
                          seed = 1999)

get_srn_results$trained_model
```

As suspected, lambda value is very small and hence regularization makes very little difference. 

```{r freeMemory, include = FALSE}
## Removing large datasets not in use anymore 
#rm(game_plays, LPM, shots, train_test_split)
```


### Random Forest 

As random forest calculation is computationally heavy and parallel computation is not yet introduced, we will make the dataset even smaller and pivot back to secondaryType explanatory factor (insted of having multiple columns for each shot type)
```{r Cleaning Data}
parsed_shots.rf <- parsed_shots %>%
  select(goal, distance, angle, secondaryType) %>%
  sample_n(size = 1000)

## Changing labels from 0 and 1 to "Goal" and "Not Goal" 
parsed_shots.rf$goal <- ifelse(test=parsed_shots.rf$goal == "1", yes = "Goal", no = "Not Goal")

## Changing the class of some columns for the commands to work
parsed_shots.rf$goal <- as.factor(parsed_shots.rf$goal)
parsed_shots.rf$secondaryType <- as.factor(parsed_shots.rf$secondaryType)

## Checking all classes are either numerical or factor
str(parsed_shots.rf)
```


```{r RandomForest, fig.show="hold", out.width="50%"}
## Making the model.
rf_model <- randomForest(goal ~ ., data = parsed_shots.rf, proximity = TRUE, ntree = 1000)

## This is to see how many trees are necessary for accurate predictions. It calculates the error rates after making of each tree. 
oob.error.data <- data.frame(
  trees=rep(1:nrow(rf_model$err.rate), times = 3),
  type=rep(c("OOB", "Goal", "Not Goal"), each=nrow(rf_model$err.rate)),
  error=c(rf_model$err.rate[,"OOB"],
          rf_model$err.rate[,"Goal"],
          rf_model$err.rate[,"Not Goal"]))

ggplot(data=oob.error.data, aes(x=trees, y=error)) +
  geom_line(aes(color=type)) +
  theme_bw() +
  ggtitle("OOB Error Rate")


## Calculating Out-of-the-bag error rates for different mtrys, which means how many variables are randomly sampled as candidates at each split. 
oob.values <- vector(length = 10)
for(i in 1:10){
  temp.rf_model <- randomForest(goal ~., data = parsed_shots.rf, mtry = i,
                                ntree = 100)
  oob.values[i] <- temp.rf_model$err.rate[nrow(temp.rf_model$err.rate), 1]
}

## Looking for the mtry that gives smallest OOB error
oob.values

## Building a MDS plot 
distance.matrix <- dist(1 - rf_model$proximity)
mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret = TRUE)
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
                       X=mds.values[,1],
                       Y = mds.values[,2],
                       Status = parsed_shots.rf$goal)

ggplot(data=mds.data, aes(x=X, y=Y, label = Sample)) +
  theme_bw() +
  geom_text(aes(color=Status)) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep = "")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep = "")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)") +
  theme_bw()
```





```{r, fig.show="hold", out.width="50%"}
varImpPlot(rf_model)

rf_model.pred <- predict(rf_model, parsed_shots.rf, select = -goal,
                         type = "class")


rf_model.pred.results <- table(rf_model.pred, parsed_shots.rf$goal)
head(rf_model.pred.results)
rf_model.pred.results
print(rf_model.accuracy <- sum(diag(rf_model.pred.results)) / sum(rf_model.pred.results))                                
```

### xGBoost

To be updated

## Sources 

Valuable sources have been:

-   <https://github.com/Dato-Futbol/xg-model/blob/master/04code_evaluate_use_models.R>
-   <https://www.thesignificantgame.com/portfolio/expected-goals-model-with-tidymodels/>
-   <https://www.datofutbol.cl/xg-model/>
-   <https://medium.com/datos-y-ciencia/una-mirada-al-soccer-analytics-usando-r-parte-iii-3bdff9cd3752>
-   <https://rstudio-pubs-static.s3.amazonaws.com/311470_f6e88d4842da46e9941cc6547405a051.html>
-   <https://soccermatics.readthedocs.io/en/latest/gallery/lesson2/plot_xGModelFit.html>
-   <https://github.com/iandragulet/xG_Model_Workflow/blob/main/xG_model_part1.ipynb>

```{r, include=FALSE}
#Next steps:
# xGboost? 
# Neural Network? 




# Sources to the bottom 
# PLOT HEADERS LARGER, IN THE MIDDLE AND BOLDED! 
# KNN analysis now that we already have the distances? 
```



















