---
title: "NHL xG"
output: 
  html_document:
    keep_md: true
    toc: true
    code_folding: show
    mathjax: local
    self_contained: false
  github_document: default
header-includes: -\usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(sportyR)
library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)
library(knitr)
require(tidymodels)
library(readr)
library(rmarkdown)
```

## Overview and Scope

The aim of this project is to build different naive Expected Goals models using LPM, logit and possibly other regression models. The data is obtained from [Kaggle](https://www.kaggle.com/datasets/martinellis/nhl-game-data). 

Valuable sources have been: 

* https://github.com/Dato-Futbol/xg-model/blob/master/04code_evaluate_use_models.R
* https://www.thesignificantgame.com/portfolio/expected-goals-model-with-tidymodels/
* https://www.datofutbol.cl/xg-model/
* https://medium.com/datos-y-ciencia/una-mirada-al-soccer-analytics-usando-r-parte-iii-3bdff9cd3752
* https://rstudio-pubs-static.s3.amazonaws.com/311470_f6e88d4842da46e9941cc6547405a051.html
* https://soccermatics.readthedocs.io/en/latest/gallery/lesson2/plot_xGModelFit.html
* https://github.com/iandragulet/xG_Model_Workflow/blob/main/xG_model_part1.ipynb


## Data Wrangling

Loading the data and creating a variable for shot data 
```{r load_data, cache = TRUE, R.options = list(width = 10000)}
game_plays <- read.csv("~/Downloads/archive/game_plays.csv")

## Making sure the data looks correct, showing first 15 samples and the response variable
str(game_plays[,c(1:15, 18)])
```

Parsing the data to include only the shots and goals
```{r shots, cache=TRUE}
shots <- game_plays %>%
  filter(event %in% c("Goal", "Shot")) %>%
  select(team_id_for, event, st_x, st_y, secondaryType)

## Creating a new column for a binary response variable if a shot was a goal or not
shots$goal <- as.integer(ifelse(shots$event == "Goal", 1,0))
```

Creating local variables for x and y cordinates
```{r cordinates}

##Taking the absolute value to have all the shots in the same horizontal axis
x <- abs(shots$st_x)
y <- shots$st_y
```

Creating functions for distance and angle. Distance is calculated by 

$c^2 = a^2 + b^2 \Rightarrow c = \sqrt{a^2 + b^2 }$

The angle to the goal is calculated as follows, 

$\tan \theta = \frac{\text{goal width } * x }{x^2 + y^2 - (\frac{\text{goal width}}{2})^2 }$

```{r functions}
distance <- function(x_pos, y_pos) {
  sqrt((89 - abs(x_pos))^2 + y_pos^2)
}

angle_theta <- function(x_pos, y_pos) {
  x_temp <- abs(89 - abs(x_pos))
    ifelse((x_temp)^2 + (y_pos)^2 >= 9, atan((6 * abs(x_temp))/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi,
    180 + atan((6 * x_temp)/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi)
}

## Making sure our range is correct, should be between 0 and 180
range(angle_theta(x,y),na.rm=TRUE)

## Adding columns for distance and angle
shots <- shots %>%
  mutate(distance = distance(shots$st_x, shots$st_y),
                    angle = angle_theta(shots$st_x, shots$st_y))
## Testing this works
head(shots)
```
## Plots

### Histograms

A couple of histograms from the data that show the distribution of shot distance and angle
```{r, Histograms, fig.show="hold", out.width="50%"}
par(mar = c(4, 4, .1, .1))
ggplot(shots, aes(x=shots$angle)) +
  geom_histogram(binwidth = 3,
                 center = 0,
                 color = "black",
                 fill = "white") +
  scale_x_continuous(limits = c(0, 180)) +
  theme_bw() +
  theme(panel.border = element_blank()) + 
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank()) + 
  xlab("Angle") + 
  ylab(element_blank()) + 
  ggtitle("Histogram of Shot Angles")

ggplot(shots, aes(x=distance)) +
    geom_histogram(binwidth = 6,
                   center = 0,
                   color = "black",
                   fill = "white") +
    scale_x_continuous(limits = c(0, 180)) +
    theme_bw() +
  theme(panel.border = element_blank()) + 
    theme(panel.grid.major = element_blank()) +
    theme(panel.grid.minor = element_blank()) + 
    xlab("Distance") + 
    ylab(element_blank()) + 
    ggtitle("Histogram of Shot Distances")
```


### Probability of a Goal Given Distance or Angle
```{r Bins, fig.show="hold", out.width="50%", cache = TRUE}

bins_distance <- aggregate(shots,
                   by=list(cut(shots$distance, seq(0,100,10))),
                   mean)

bins_angle <- aggregate(shots,
                  by=list(cut(shots$angle, seq(0,180,10))),
                  mean)

## Changing the first column to numeric values so that ggplot geom_smooth works
bins_distance$Group.1 <- as.numeric(bins_distance$Group.1)
bins_angle$Group.1 <- as.numeric(bins_angle$Group.1)

angles <- as.character(seq(0, 180, 10))
distances <- as.character(seq(0, 90, 10))


ggplot(bins_distance, aes(x= bins_distance$Group.1, y =  bins_distance$goal)) +
                          geom_point() +
                          geom_line() +
                          theme_bw() + 
                          xlab("Distance to goal (Feet)") +
                          ylab("Probability of Goal") + 
                          scale_x_discrete(limits = distances)
                          
ggplot () + aes(x= bins_angle$Group.1, y =  bins_angle$goal) +
  geom_point() +
  geom_smooth(method=lm, se = F) + 
  theme_bw() +
  xlab("Angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Probability of Goal Given the Distance") +
  scale_x_discrete(limits = angles)
```
In the distance to goal there's an interesting fact: probability of goal increases with distance. This is likely due to the fact that usually shots from very far away are shot due to empty goal: hence it scewing the data. In angle to goal there's no notable surprises. 


This chunk is saved for possible future uses. To ensure unbiasedness, training data has to be separated from the actual testing data. With the following commands the sample is randomized and 70% of it would be used for the training data. 
```{r training_data}
#train_test_split <- initial_split(data = shots, prop = 0.7)

#train_data <- train_test_split %>%
#  training()
#test_data <- train_test_split %>%
#  testing()
```

```{r LPM}
LPM <- lm(goal ~ distance + angle, data = shots)
summary(LPM)

LPM_distance <- as.numeric(LPM$coefficients["distance"])
LPM_angle <- as.numeric(LPM$coefficients["angle"])
LPM_intercept <- as.numeric(LPM$coefficients["(Intercept)"])
LPM_manual <- LPM_intercept + LPM_distance * shots$distance + LPM_angle * shots$angle
```
In the plot below, the the main downside of LPM model becomes apparent: results are not bound [0,1]. 
```{r LPM Plot}
ggplot(data = LPM, mapping=aes(x=angle, y = goal)) +
  geom_point() + geom_smooth(method = "lm", se = F)
```

Because of this and various other reasons, logit is used. 

```{r logit}
logit <- glm(goal ~ distance + angle,
             family = binomial(link = 'logit'),
             data = shots)

summary(logit)

logit_distance <- as.numeric(logit$coefficients["distance"])
logit_angle <- as.numeric(logit$coefficients["angle"])
logit_intercept <- as.numeric(logit$coefficients["(Intercept)"])
logit_value <- 1/(1+exp(logit_intercept + logit_distance * shots$distance + logit_angle * shots$angle))

```
In a logit model, the probability of an event is given by 

$P = \frac{1}{1 + - exp^{-{(\beta_0 + \beta_1 x_1 \beta_2 x_2 + …)}}}$

```{r Logit Plots, fig.show="hold", out.width="50%"}
ggplot(logit, aes(x=distance, y =goal)) +
  geom_point() + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  scale_x_reverse() +
  theme_bw() +
  xlab("Distance to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Distance as an explanatory variable") 

ggplot(logit, aes(x=angle, y =goal)) +
  geom_point() + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  theme_bw() +
  xlab("Angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Angle as an explanatory variable") 
```
From graphs above, it becomes visually clear that angle is a way more important factor affecting if a shot is a goal or not. 

```{r Heatmap LPM}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$xg <- LPM_intercept + distance(artificial_shots$location_x,artificial_shots$location_y) * LPM_distance + angle_theta(artificial_shots$location_x, artificial_shots$location_y) * LPM_angle

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg, alpha = 1)) +
  scale_color_gradient2(low = "white", mid="red", midpoint = 0.55, high ="darkred",
                       scales::rescale(c(0.9,0.1))) +
                       ggtitle("LPM Heatmap")
```


```{r Heatmap logit}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$xg_logit <- 1 / (1 + exp(-logit_intercept - distance(artificial_shots$location_x,artificial_shots$location_y) * logit_distance - angle_theta(artificial_shots$location_x, artificial_shots$location_y) * logit_angle))


geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg_logit, alpha = 0.1)) +
  scale_color_gradient(low = "white", high ="red",
                       scales::rescale(c(0.1,0.9))) + 
                        ggtitle("Heatmap Logit")
                        
```

















