---
title: "NHL xG"
output:
  github_document:
    toc: true
  #toc-title: "Table of Contents"
  code_folding: show
#  toc_depth: 2
#    code_folding: show
#    mathjax: local
#    self_contained: false
#header-includes: -\usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2023)
library(sportyR)
library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)
library(knitr)
require(tidymodels)
library(readr)
library(rmarkdown)
library(pROC)
library(randomForest)
library(cowplot)
library(usethis)
library(parallel)
library(Hmisc)
```

## Overview and Scope

The aim of this project is to build different naive Expected Goals models using LPM, logit and possibly other regression models. The data is obtained from [Kaggle](https://www.kaggle.com/datasets/martinellis/nhl-game-data).

Valuable sources have been:

-   <https://github.com/Dato-Futbol/xg-model/blob/master/04code_evaluate_use_models.R>
-   <https://www.thesignificantgame.com/portfolio/expected-goals-model-with-tidymodels/>
-   <https://www.datofutbol.cl/xg-model/>
-   <https://medium.com/datos-y-ciencia/una-mirada-al-soccer-analytics-usando-r-parte-iii-3bdff9cd3752>
-   <https://rstudio-pubs-static.s3.amazonaws.com/311470_f6e88d4842da46e9941cc6547405a051.html>
-   <https://soccermatics.readthedocs.io/en/latest/gallery/lesson2/plot_xGModelFit.html>
-   <https://github.com/iandragulet/xG_Model_Workflow/blob/main/xG_model_part1.ipynb>

## Data Wrangling

Loading the data and creating a variable for shot data

```{r load_data, cache = TRUE, R.options = list(width = 10000)}
game_plays <- read.csv("~/Downloads/archive/game_plays.csv")

## Making sure the data looks correct, showing first 18 samples and the response variable
str(game_plays)
```

Parsing the data to include only the shots and goals

```{r shots, cache=TRUE}
shots <- na.omit(game_plays) %>%
  filter(event %in% c("Goal", "Shot")) %>%
  select(team_id_for, event, st_x, st_y, secondaryType)

## Creating a new column for a binary response variable if a shot was a goal or not
shots$goal <- as.integer(ifelse(shots$event == "Goal", 1,0))
```

Creating local variables for x and y cordinates

```{r cordinates}

##Taking the absolute value to have all the shots in the same horizontal axis
x <- abs(shots$st_x)
y <- shots$st_y
```

Creating functions for distance and angle. Distance is calculated by

$c^2 = a^2 + b^2 \Rightarrow c = \sqrt{a^2 + b^2 }$

The angle to the goal is calculated as follows,

$\tan \theta = \frac{\text{goal width } * x }{x^2 + y^2 - (\frac{\text{goal width}}{2})^2 }$

```{r functions}
distance <- function(x_pos, y_pos) {
  sqrt((89 - abs(x_pos))^2 + y_pos^2)
}

angle_theta <- function(x_pos, y_pos) {
  x_temp <- abs(89 - abs(x_pos))
    ifelse((x_temp)^2 + (y_pos)^2 >= 9, atan((6 * abs(x_temp))/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi,
    180 + atan((6 * x_temp)/(x_temp^2 + y_pos^2 - 3^2)) * 180 / pi)
}

## Making sure our range is correct, should be between 0 and 180
range(angle_theta(x,y),na.rm=TRUE)

## adding columns for distance and angle
shots <- shots %>%
  mutate(distance = distance(shots$st_x, shots$st_y),
                    angle = angle_theta(shots$st_x, shots$st_y))
## Testing this works
head(shots)
```

## Data Exploration

### Histograms

a couple of histograms from the data that show the distribution of shot distance and angle

```{r, Histograms, fig.show="hold", out.width="50%", class.source = 'fold-hide'}
par(mar = c(4, 4, .1, .1))
ggplot(shots, aes(x=shots$angle)) +
  geom_histogram(binwidth = 3,
                 center = 0,
                 color = "black",
                 fill = "white") +
  scale_x_continuous(limits = c(0, 180)) +
  theme_bw() +
  theme(panel.border = element_blank()) + 
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank()) + 
  xlab("angles") + 
  ylab(element_blank()) + 
  ggtitle("Histogram of Shot angles")

ggplot(shots, aes(x=distance)) +
    geom_histogram(binwidth = 6,
                   center = 0,
                   color = "black",
                   fill = "white") +
    scale_x_continuous(limits = c(0, 180)) +
    theme_bw() +
  theme(panel.border = element_blank()) + 
    theme(panel.grid.major = element_blank()) +
    theme(panel.grid.minor = element_blank()) + 
    xlab("Distance") + 
    ylab(element_blank()) + 
    ggtitle("Histogram of Shot Distances")
```

### Probability of a Goal Given Distance or angle

```{r Bins, fig.show="hold", out.width="50%", cache = TRUE}

bins_distance <- aggregate(shots,
                   by=list(cut(shots$distance, seq(0,100,10))),
                   mean)

bins_angle <- aggregate(shots,
                  by=list(cut(shots$angle, seq(0,180,10))),
                  mean)

## Changing the first column to numeric values so that ggplot geom_smooth works
bins_distance$Group.1 <- as.numeric(bins_distance$Group.1)
bins_angle$Group.1 <- as.numeric(bins_angle$Group.1)

angles <- as.character(seq(0, 180, 10))
distances <- as.character(seq(0, 90, 10))


ggplot(bins_distance, aes(x= bins_distance$Group.1, y =  bins_distance$goal)) +
                          geom_point() +
                          geom_line() +
                          theme_bw() + 
                          xlab("Distance to goal (Feet)") +
                          ylab("Probability of Goal") + 
                          scale_x_discrete(limits = distances) +
                          ggtitle("Probability of Goal Given the Distance")
                          
ggplot () + aes(x= bins_angle$Group.1, y =  bins_angle$goal) +
  geom_point() +
  geom_smooth(method=lm, se = F) + 
  theme_bw() +
  xlab("angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Probability of Goal Given the angle") +
  scale_x_discrete(limits = angles)
```

In the distance to goal there's an interesting fact: probability of goal increases with distance. This is likely due to the fact that usually shots from very far away are shot due to empty goal: hence it scewing the data. In angle to goal there's no notable surprises.


### Shot Plots
```{r shotPlots, fig.show="hold", out.width="50%"}
#Shot maps for randomly samples 2500 shots 
shots_parsed <- shots %>%
  subset(st_x < 88) %>%
  sample_n(2500)

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = shots_parsed$st_y, y = shots_parsed$st_x, col = shots_parsed$goal, size = shots$parsed$goal, alpha = 0.5)) +
  scale_color_binned(low ="red", high = "darkgreen")

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_density2d_filled(aes(x = shots_parsed$st_y, y = shots_parsed$st_x, alpha = 0.5),
                        contour_var = "ndensity",
                        breaks = seq(0.1, 1.0, length.out = 10))
```

## Basic Regression Models

### Linear Probability Model

```{r LPM}
LPM <- lm(goal ~ distance + angle, data = shots)
summary(LPM)
```

In the plot below, the the main downside of LPM model becomes apparent: results are not bound [0,1].

```{r LPM Plot}
ggplot(data = LPM, mapping=aes(x=angle, y = goal)) +
  geom_point() + geom_smooth(method = "lm", se = F) +
  theme_bw()
```

```{r Heatmap LPM}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$xg <- predict(LPM, artificial_shots, type = "response")

geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg, alpha = 1)) +
  scale_color_gradient2(low = "white", mid="red", midpoint = 0.55, high ="darkred",
                       scales::rescale(c(0.9,0.1)))
```

### Logit Part 1
Due to the significant downsides of LPM, logistic regression is henceforth used. 

```{r logit}
logit <- glm(goal ~ distance + angle,
             family = binomial(link = 'logit'),
             data = shots)


```

In a logit model, the probability of an event is given by

$P = \frac{1}{1 + - exp^{-{(\beta_0 + \beta_1 x_1 \beta_2 x_2 + …)}}}$

```{r Logit Plots, fig.show="hold", out.width="50%"}
ggplot(logit, aes(x=distance, y =goal)) +
  geom_point() + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  scale_x_reverse() +
  theme_bw() +
  xlab("Distance to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("Distance as an explanatory variable") 

ggplot(logit, aes(x=angle, y =goal)) +
  geom_point() + geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = F) +
  theme_bw() +
  xlab("angle to Goal") +
  ylab("Probability of Goal") + 
  ggtitle("angle as an explanatory variable") 
```

From graphs above, it becomes visually clear that angle is a way more important factor affecting if a shot is a goal or not. To test whether we could improve explanatory power of distance, we add a quadratic form of it as an extra variable.

```{r Quadratics}
shots$distance_sq <- shots$distance^2

logit.2 <- glm(goal ~ distance + distance_sq + angle,
               family = binomial(link = 'logit'),
               data = shots)

summary(logit.2)
logit.2_coef <- logit.2$coefficients
logit.2_distance <- logit.2_coef["distance"]
logit.2_distance_sq <- logit.2_coef["distance_sq"]
logit.2_intercept <- logit.2_coef["(Intercept)"]


b <- data.frame(c(seq(0,150,.1)))

a <- (1 / (1 + exp(-logit.2_distance * b - logit.2_distance_sq * b - logit.2_intercept)))
a.2 <- (1 / (1 + exp(-logit.2_distance * b - logit.2_intercept)))

c <- cbind(a, a.2, b)


head(c)
colnames(c) <- c("a", "a.2", "b")

ggplot(c, aes(x=b,y=a.2)) +
  geom_point(size = 2, col = "darkgreen") + 
  geom_point(aes(y=a.2), size = 0.1, col = "darkred") +
  theme_bw() + 
  ggtitle("Comparing Distance Variables with and without quadratic term") +
  xlab("Distance to Goal") +
  ylab("Probability of a Goal")
```

```{r, include = FALSE}
#Overwriting logit.2
logit.2 <- glm(goal ~ distance + angle,
               family = binomial(link = 'logit'),
               data = shots)
```


As we can see, quadratic's effect is just taking into account the 'long-shot bias', which is caused by empty-net goals.



```{r Heatmap logit, fig.show="hold"}
artificial_shots <- crossing(location_x = seq(30, 88, by = 1), location_y = seq(-37, 37, by = 1))

artificial_shots$distance <- distance(artificial_shots$location_x, artificial_shots$location_y)
artificial_shots$angle <- angle_theta(artificial_shots$location_x, artificial_shots$location_y)
#artificial_shots$xg_logit <- 1 / (1 + exp(-logit_intercept - distance(artificial_shots$location_x,artificial_shots$location_y) * logit_distance - angle_theta(artificial_shots$location_x, artificial_shots$location_y) * logit_angle))
artificial_shots


artificial_shots$xg_logit <- predict(logit, artificial_shots, type = "response")


geom_hockey(league = "NHL", rotation = 90, display_range = "ozone") +
  geom_point(aes(x = artificial_shots$location_y, y = artificial_shots$location_x, col = artificial_shots$xg_logit, alpha = 0.1)) +
  scale_color_gradient(low = "white", high ="red",
                       scales::rescale(c(0.1,0.9))) 
```


## Other Regression and Classification Models 

### Introduction and adding variables 

Note: this part is more experimental and is very prone to mistakes. 

Henceforth we will be comparing the effectivness of the models, hence the data will be split into training- and testing data. Here 70% of the full sample is used for training and the remaining 30% for testing. This ensured unbiasedness when testing the models.

```{r training_data}

parsed_shots <- shots %>%
  select(goal, distance, angle, secondaryType)
train_test_split <- initial_split(data = parsed_shots, prop = 0.7)

train_data <- train_test_split %>%
  training()
test_data <- train_test_split %>%
  testing()
```

Let's broaden our analysis by adding an other explanatory factor to the regression: shot type. We have the following options:

```{r}
unique(shots$secondaryType)
```

### Logit Part 2

Lets add this to the regression and see how the coefficients for distance and angle change.

```{r adding secondaryType}
logit.3 <- glm(goal ~ distance + angle + secondaryType,
             family = binomial(link = 'logit'),
             data = train_data)

logit_pred <- predict(logit, test_data, type = "response")
logit.2_pred <- predict(logit.2, test_data, type = "response")

logit.3_pred <- predict(logit.3, test_data, type = "response")
table(shots$goal)

head(logit.3_pred)
sum(logit.3_pred > 0.01, na.rm=TRUE)


#Comparing Logit models 
roc.test(roc(test_data$goal, logit.3_pred), roc(test_data$goal, logit.2_pred))

#Comparing better Logit model to the LPM model
roc.test(roc(test_data$goal, logit.3_pred), roc(test_data$goal, logit_pred))
par(pty = "s")

logit.3_roc <- roc(test_data$goal, logit.3_pred, plot = TRUE, print.auc = TRUE, col = "darkred",
  legacy.axes = TRUE, percent = TRUE, xlab = "False Positive Percentage",
  ylab = "True Positive Percentage") 

#Observing the optimal threshold level and the corresponding specificity and sensitivity levels
logit.3_threshold <- coords(logit.3_roc, "best", "threshold")
logit.3_threshold

#Creating Confusion Matrix
logit.3_conf <- table(logit.3_pred>=logit.3_threshold$threshold, test_data$goal)
sum(diag(logit.3_conf))/sum(logit.3_conf)*100

```

```{r freeMemory, include = FALSE}
## Removing large datasets not in use anymore 
rm(game_plays, LPM, shots, train_test_split)

```


### Random Forest 
```{r Cleaning Data}
## Making the train_data smaller due to memory issues 
train_data <- sample_n(train_data, size = 1000)

## Changing labels from 0 and 1 to "Goal" and "Not Goal" 
train_data$goal <- ifelse(test=train_data$goal == "1", yes = "Goal", no = "Not Goal")
test_data$goal <- ifelse(test=test_data$goal == 0, yes = "Not Goal", no = "Goal")

## Changing the class of some columns for the commands to work
train_data$goal <- as.factor(train_data$goal)
train_data$secondaryType <- as.factor(train_data$secondaryType)
test_data$goal <- as.factor(test_data$goal)
test_data$secondaryType <- as.factor(test_data$secondaryType)

## Checking all classes are either numerical or factor
str(train_data)

## Calculating the amount of NA's 
sum(is.na(train_data))
```
As we can see, we have 15 NA's in the 'Angle' Column. This is fixed by na.action = na.roughfix, which fills NA's with column median. Because of the huge sample size, more advanced imputations are not needed. 


```{r RandomForest}
## Making the model 
#set.seed(2023)
rf_model <- randomForest(goal ~ ., data = train_data, proximity = TRUE, na.action = na.roughfix, ntree = 1000)

## This is to see how many trees are necessary for accurate predictions. It calculates the error rates after making of each tree. 
oob.error.data <- data.frame(
  trees=rep(1:nrow(rf_model$err.rate), times = 3),
  type=rep(c("OOB", "Goal", "Not Goal"), each=nrow(rf_model$err.rate)),
  error=c(rf_model$err.rate[,"OOB"],
          rf_model$err.rate[,"Goal"],
          rf_model$err.rate[,"Not Goal"]))
ggplot(data=oob.error.data, aes(x=trees, y=error)) +
  geom_line(aes(color=type))


## Calculating Out-of-the-bag error rates for different mtrys, which means how many variables are randomly sampled as candidates at each split. 
oob.values <- vector(length = 10)
for(i in 1:10){
  temp.rf_model <- randomForest(goal ~., data = train_data, mtry = i,
                                ntree = 100)
  oob.values[i] <- temp.rf_model$err.rate[nrow(temp.rf_model$err.rate), 1]
}

## Looking for the mtry that gives smallest OOB error
oob.values

## Building a MDS plot 
distance.matrix <- dist(1 - rf_model$proximity)
mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret = TRUE)
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
                       X=mds.values[,1],
                       Y = mds.values[,2],
                       Status = train_data$goal)
head(mds.values)

ggplot(data=mds.data, aes(x=X, y=Y, label = Sample)) +
  theme_bw() +
  geom_text(aes(color=Status)) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep = "")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep = "")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")
```






```{r, include=FALSE}
#Next steps:
#-figure out if the model is broken or just very bad 
#-Gradient Boosting?
#xGboost? 



#Fixing logit plots in the beginning 
```



















